Report on the Second Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE2)

### 2015-06-01 WSSSPE2_report/WSSSPE2_review.tex

It was suggested that these concrete things should be small things so that the group members were likely to do them.

An example was the Architecture of Open Source Applications book~\cite{aosa}, containing contributed chapters on open source software; the group wondering if it could do the same thing for testing?

The group asked if it could produce a 100-line example of how to test neutron transport (or other specific scientific examples) targeted at a sophomore? This would be a demonstration of ``How do I do this right?''

Another question was ``How do I test so that I know that the code is not the source of my problems?''
A challenge is stripping things down to an appropriate level.

Testing for infrastructure is easier than testing for models. \katznote{to delete? - don't see how it fits}

The group referred to Lorena Barba's ``12 Steps to Navier Stokes~\cite{12steps},'' an online CFD lesson repository created as a set of iPython notebooks.

@@

Distinction between 100 lines of input files vs. 100 lines of source code \katznote{to delete? - don't see how it fits}

@@

One general rule that was suggested is that projects shouldn't accept patches that don't have associated tests.  This led to discussion about the fact that setting up infrastructure to test well is difficult.  The group would like a standard way that runs out of the box.  Jenkins was suggested as
a solution, but setting it up is considered difficult for most scientists.

Group members also comments that people who've worked with a testing framework once (even if they have to be forced to do so) won't work without it afterward.  But how to get them to do this the first time is not clear.

Two important pieces of the barrier are a) picking/implementing testing frameworks (a technical barrier) and b) deciding what are the actual tests that I need to write?  A possible method to solve the latter is to ask students ``how is what you're doing different from typing random keys on the keyboard?'' and then, turn their answer into the concept of tests.

Another point of discussion is that many scientists are not aware that ``testing the code'' and ``testing the science'' are separate.  A related question that came up is ``How do you know that your method produces a result that is `close enough'?''

@@

Tom Clune suggested it would be useful to have a piece of code that is badly engineered and lacks unit tests so that he can add the unit tests and show people what it should look like.  However, such a collaboration must be with someone that either wrote the code or is intimately familiar with the topic.  The exercise will be to extract unit-tests through reverse-engineering {\b and} questioning the developer/expert.
Others suggested that the code does not have to be ``badly engineered'', but is either representative or is believed to be impenetrable from a unit-testing perspective.

The group discussed how they learned testing.  One attendee learned from a TDD tutorial at a Best Practices conference. The examples were {\b not} relevant, partly since he was working with Fortran code in his job, so the remainder of his learning was in the form of proving it could be done with Fortran.  Another attendee said that most scientists have broken out pieces of software from a larger application and done some checking in an informal way.  The difference in his work is that he now knows about frameworks that let him do that systematically.
He realized he was always building on top of other people’s software, and eventually realized that he should adopt those good software habits.  He also commented that
most of those tests are ephemeral.  They live long enough to give the author of the test confidence in code, then the author moves on; the tests are not handed on and made available to next user.

Attendees joked that they started testing when their skepticism got bigger than their arrogance, or when their guilt got bigger than my arrogance.

>>>>

The consensus of the group was that small, concrete actions would have
a greater chance of proper follow-through by mpmebers of the group.
An example of this approach was the Architecture of Open Source
Applications book~\cite{aosa}, containing contributed chapters on open
source software.  
The group contemplated creating an analogous book/web-resource for
testing of scientific software.

Several members of the discussion indicated that while they valued the
concpet software testing in theory, they were unaware how to test
certain kinds of software.  
Or, more importantly, they lacked relevant examples of software
testing that would be suitable for new members to their team.
The group asked if it could produce a 100-line example of how to test
neutron transport (or other specific scientific examples) targeted at
a sophomore? This would be a demonstration of ``How do I do this
right?''

Another question was ``How do I test so that I know that the code is
not the source of my problems?''  A common challenge is that of
stripping things down to an appropriate level for tests.  The group
recognized that the difficulty is different for so-called software
infrastructure than for numerical/scientific layers of software
applications.

The group referred to Lorena Barba's ``12 Steps to Navier
Stokes~\cite{12steps},'' an online CFD lesson repository created as a
set of iPython notebooks.  \clunenote{I'm not sure of the context
  here, except possibly as an analog for something under discussion.}

@@

Distinction between 100 lines of input files vs. 100 lines of source code \katznote{to delete? - don't see how it fits}
\clunenote{I no longer recall the context for this statement; does not seem important to me.}

@@

One general rule that was suggested is that projects should not accept
software patches that don't have associated tests.  
This led to discussion about the fact that setting up appropriate 
infrastructure for testing can be difficult.
The group would like a standard way that runs out of the box.
Jenkins was suggested as a common open-source solution, but initial
configuration is considered to be challenging for typical scientists.

Several members of the group commented that once a developer has been
exposed to a testing framework in a nontrivial fashion, he/she will
subsequently insist upon using a framework for further work.
The acknowledged challenge was how to create that crucial first exposure
to such frameworks.

Two important pieces of the barrier are a) picking/implementing testing frameworks (a technical barrier) and b) deciding what are the actual tests that I need to write?  A possible method to solve the latter is to ask students ``how is what you are doing different from typing random keys on the keyboard?'' and then, turn their answer into the concept of tests.  I.e. expand the process of creating code from one that is solely about ``how?'' into one that also include ``why?''.

Another point of discussion is that many scientists are not aware that ``testing the code'' and ``testing the science'' are distinct issues.  A related question that came up is ``How do you know that your method produces a result that is `close enough'?''

@@

Tom Clune suggested that a useful demonstration could be generate by
beginning with a (smallish) piece of badly engineered code that lacks
appropriate unit tests.   
The code would then be refactored into a more presentable form once
unit test were created/derived.
However, such a collaboration must be with someone that either wrote
the code or is intimately familiar with the subject matter.  The exercise would
be to extract unit-tests through reverse-engineering {\b and}
questioning the developer/expert.  Others suggested that the code does
not have to be ``badly engineered'', but is either representative or
is believed to be impenetrable from a unit-testing perspective.

The group discussed how they learned testing.  One attendee learned from a tutorial on test driven development (TDD) at a Software Developers Best Practices conference. The examples were {\b not} relevant, partly since he was working with Fortran code in his job, so the remainder of his learning was in the form of proving the process could be applied with Fortran.  Another attendee said that most scientists have broken out pieces of software from a larger application and done some checking in an informal way.  The difference in his work is that he now knows about frameworks that let him do that systematically.  He realized he was always building on top of other people’s software, and eventually realized that he should adopt those good software habits.  He also commented that
most of those initial checks were ephemeral.  They live long enough to give the author of the test confidence in code, then the author moves on; the checks  are not handed on and made available to next user as permanent tests.

Attendees joked that they started testing when their skepticism/guilt became larger than their arrogance.
